### 2020-10-09 conclusion.tex
Cleaned up BibTeX files. Removed 'seminal'.

In the standard typed $\pi$-calculus a seminal line of work is by Kobayashi and his original type-theoretic approach to deadlock-freedom \cite{K98}, where priorities are abstract tags defined over a partially ordered set. Later on these abstract tags were simplified to natural numbers, and pairs of obligations and capabilities were used in the type system for deadlock freedom \cite{K02,K06}. This allowed more $\pi$-calculus processes to be typed. Following, Kobayashi's line of work, Padovani \cite{P13} adapted obligation/capability pairs to session types, and later on he further simplified them to a single priority for linear $\pi$-calculus \cite{P14}. As discussed by the author, by using the encoding of session types into linear types \cite{Koba07,DGS12,D14,Dardha16}, the deadlock freedom technique of the linear $\pi$-calculus with priorities can be transferred onto the $\pi$-calculus with session types.

The seminal work on progress for session-typed $\pi$-calculus, by Dezani \emph{et al.} \cite{DMYD06}, guarantees progress by allowing only one active session at a time. In \cite{DLY07} the authors introduce a partial order on channels in line with Kobayashi's work \cite{K98}. Carbone and Debois \cite{CD10} define progress for session typed $\pi$-calculus in terms of a \emph{catalyser} used to provide a missing counterpart to a process, thus guaranteeing progress.
Carbone \emph{et al.} \cite{CDM14} studied further the use of catalysers and showed that progress is a compositional form of lock-freedom. As in \cite{P14} the authors show that by using the encoding of session types \cite{DGS12} and Kobayashi's obligations/capabilities, we can obtain progress in for session types. Vieira and Vasconcelos \cite{VieiraV13} used single priorities and an abstract partial order to guarantee deadlock freedom in session-typed $\pi$-calculus.

\paragraph{On Curry-Howard correspondences}
The Curry-Howard correspondence between logic and types laid the foundation for functional programming \cite{Wadler15}. With the rise of linear logic \cite{Girard87} a Curry-Howard correspondence was established between linear logic and the linear $\pi$-calculus \cite{Abramsky94,BellinS94}. With the rise of session types, a new correspondence between linear logic and the $\pi$-calculus was proposed first by Caires and Pfenning \cite{CP10}, for dual intuitionistic linear logic (DILL), and later by Wadler \cite{wadler2012} for classical linear logic (CLL). The underlying language in both correspondences satisfy deadlock freedom by design, by forbidding cyclic connections via typing. Dardha and Gay follow Padovani's work \cite{P14} and use priorities in CP to define a new calculus which is deadlock-free by design, based on classical linear logic \cite{Girard87}  and the Curry-Howard correspondence with session types \cite{wadler2012}. Dardha and P\'{e}rez \cite{DardhaP15} compare Kobayashi-style typing and CLL typing, and prove that CLL corresponds to Kobayashi's system with the restriction that only single cuts are allowed. Balzer \emph{et al.} study sharing \cite{BalzerP17} and later on deadlock freedom \cite{BalzerTP19} in session-typed $\pi$-calculus based on DILL. In \cite{BalzerP17} types are either linear or shared with modal operators connecting them, which allows sharing, but not deadlock freedom. Later on in \cite{BalzerTP19} types are enhanced with information regarding the order of acquiring resources, which guarantees deadlock freedom.

\paragraph{Conclusion and Future Work}
In this paper we presented Priority GV, which is a functional language with session types. We used priorities in session types, following Priority CP \cite{dardha2018} to guarantee deadlock freedom statically via our priority-based type system. We proved the progress property of PGV in \Cref{lem:pgv-closed-progress-confs}. We also presented an updated version of PCP, where we moved from commuting conversions to structural congruence in the reduction relation, thus remaining faithful to the reduction relation of the $\pi$-calculus. Finally, we presented an encoding of Priority CP to Pririty GV in \Cref{fig:pcp-to-pgv}. For this translation we proved that it is sound with respect to typing (\Cref{{lem:pcp-to-pgv-terms-preservation}}, \Cref{thm:pcp-to-pgv-confs-preservation}) and operational semantics, given by operational correspondence soundness (\Cref{thm:pcp-to-pgv-operational-correspondence-soundness}) and completeness (\Cref{thm:pcp-to-pgv-operational-correspondence-completeness}).

>>>>

In the standard typed $\pi$-calculus a foundational line of work is by Kobayashi and his original type-theoretic approach to deadlock-freedom~\cite{kobayashi98}, where priorities are abstract tags defined over a partially ordered set. Later on these abstract tags were simplified to natural numbers, and pairs of obligations and capabilities were used in the type system for deadlock freedom~\cite{kobayashi02,kobayashi06}. This allowed more $\pi$-calculus processes to be typed. Following, Kobayashi's line of work, Padovani~\cite{padovani13} adapted obligation/capability pairs to session types, and later on he further simplified them to a single priority for linear $\pi$-calculus~\cite{padovani14}. As discussed by the author, by using the encoding of session types into linear types~\cite{kobayashi07,dardhagiachino12,dardha14beat,dardha16}, the deadlock freedom technique of the linear $\pi$-calculus with priorities can be transferred onto the $\pi$-calculus with session types.

The groundbreaking work on progress for session-typed $\pi$-calculus, by Dezani \emph{et al.}~\cite{dezani-ciancaglinimostrous06}, guarantees progress by allowing only one active session at a time. In~\cite{dezani-ciancagliniliguoro09progress} the authors introduce a partial order on channels in line with Kobayashi's work~\cite{kobayashi98}. Carbone and Debois~\cite{carbonedebois10} define progress for session typed $\pi$-calculus in terms of a \emph{catalyser} used to provide a missing counterpart to a process, thus guaranteeing progress.
Carbone \emph{et al.}~\cite{carbonedardha14} studied further the use of catalysers and showed that progress is a compositional form of lock-freedom. As in~\cite{padovani14} the authors show that by using the encoding of session types~\cite{dardhagiachino12} and Kobayashi's obligations/capabilities, we can obtain progress in for session types. Vieira and Vasconcelos~\cite{vieiravasconcelos13} used single priorities and an abstract partial order to guarantee deadlock freedom in session-typed $\pi$-calculus.

\paragraph{On Curry-Howard correspondences}
The Curry-Howard correspondence between logic and types laid the foundation for functional programming~\cite{wadler12}. With the rise of linear logic~\cite{girard87} a Curry-Howard correspondence was established between linear logic and the linear $\pi$-calculus~\cite{abramsky94,bellinscott94}. With the rise of session types, a new correspondence between linear logic and the $\pi$-calculus was proposed first by Caires and Pfenning~\cite{cairespfenning10}, for dual intuitionistic linear logic (DILL), and later by Wadler~\cite{wadler12} for classical linear logic (CLL). The underlying language in both correspondences satisfy deadlock freedom by design, by forbidding cyclic connections via typing. Dardha and Gay follow Padovani's work~\cite{padovani14} and use priorities in CP to define a new calculus which is deadlock-free by design, based on classical linear logic~\cite{girard87}  and the Curry-Howard correspondence with session types~\cite{wadler12}. Dardha and P\'{e}rez~\cite{dardhaperez15} compare Kobayashi-style typing and CLL typing, and prove that CLL corresponds to Kobayashi's system with the restriction that only single cuts are allowed. Balzer \emph{et al.} study sharing~\cite{balzerpfenning17} and later on deadlock freedom~\cite{balzertoninho19} in session-typed $\pi$-calculus based on DILL. In~\cite{balzerpfenning17} types are either linear or shared with modal operators connecting them, which allows sharing, but not deadlock freedom. Later on in~\cite{balzertoninho19} types are enhanced with information regarding the order of acquiring resources, which guarantees deadlock freedom.

\paragraph{Conclusion and Future Work}
In this paper we presented Priority GV, which is a functional language with session types. We used priorities in session types, following Priority CP~\cite{dardhagay18} to guarantee deadlock freedom statically via our priority-based type system. We proved the progress property of PGV in \Cref{lem:pgv-closed-progress-confs}. We also presented an updated version of PCP, where we moved from commuting conversions to structural congruence in the reduction relation, thus remaining faithful to the reduction relation of the $\pi$-calculus. Finally, we presented an encoding of Priority CP to Pririty GV in \Cref{fig:pcp-to-pgv}. For this translation we proved that it is sound with respect to typing (\Cref{{lem:pcp-to-pgv-terms-preservation}}, \Cref{thm:pcp-to-pgv-confs-preservation}) and operational semantics, given by operational correspondence soundness (\Cref{thm:pcp-to-pgv-operational-correspondence-soundness}) and completeness (\Cref{thm:pcp-to-pgv-operational-correspondence-completeness}).

### 2020-10-09 introduction.tex
Cleaned up BibTeX files. Removed 'seminal'.

\begin{itemize}
\item GV is a session-typed functional language.
\item
  The benefits of using a functional language as opposed to a process calculus are that:
  \begin{itemize}
  \item 
    it supports higher-order functions (and therefore abstraction) not usually present in process calculi; and
  \item
    it allows us to derive extensions of the communication fragment of the language via well-understood extensions of the functional fragment, i.e., deriving internal/external choice from sum types.
  \end{itemize}
\item
  The benefits of using GV over other session-typed functional languages are that:
  \begin{itemize}
  \item
    GV has strong ties to linear logic~\cite{wadler15}.
  \end{itemize}
\item
  Historically, calculi in the GV family~\cite{wadler15,lindley15} have achieved deadlock freedom via a syntactic restriction, i.e., by combining channel creation and thread creation into a single operation, called fork. Unfortunately, this is overly restrictive, as it limits the possible communication structures to trees. The calculus GV is based on, on the other hand, did not satisfy deadlock freedom~\cite{GAY2019}.
\end{itemize}

\begin{itemize}
\item talking about what is GV and its "benefits"
\item next, talking about deadlock freedom property, the benefits and the fact that the original GVs do not satisfy DF
\item in the meantime, in a concurrent setting based on pi-calculus there is a curry-howard correspondence of logic and session typed pi, where DF holds
\item talk about CP here
\item \cite{dardha2018} brought the research fwd and added priorities a la Kobayashi and Padovani to CP, producing PCP
\item following that lead, we present PGV
\item running example of the scheduler in PGV (hopefully)
\item translation of PCP to PGV and what it means to have that.
\item a contribution paragraph with the following
  \begin{itemize}
  \item PGV and what are the benefits
  \item updated PCP
  \item encoding PCP to PGV
  \item technical results to be referred to explicitly
  \end{itemize}
\end{itemize}

>>>>

\begin{itemize}
\item GV is a session-typed functional language.
\item
  The benefits of using a functional language as opposed to a process calculus are that:
  \begin{itemize}
  \item 
    it supports higher-order functions (and therefore abstraction) not usually present in process calculi; and
  \item
    it allows us to derive extensions of the communication fragment of the language via well-understood extensions of the functional fragment, i.e., deriving internal/external choice from sum types.
  \end{itemize}
\item
  The benefits of using GV over other session-typed functional languages are that:
  \begin{itemize}
  \item
    GV has strong ties to linear logic~\cite{wadler12}.
  \end{itemize}
\item
  Historically, calculi in the GV family~\cite{wadler12,lindleymorris15} have achieved deadlock freedom via a syntactic restriction, i.e., by combining channel creation and thread creation into a single operation, called fork. Unfortunately, this is overly restrictive, as it limits the possible communication structures to trees. The calculus GV is based on, on the other hand, did not satisfy deadlock freedom~\cite{gayvasconcelos10}.
\end{itemize}

\begin{itemize}
\item talking about what is GV and its "benefits"
\item next, talking about deadlock freedom property, the benefits and the fact that the original GVs do not satisfy DF
\item in the meantime, in a concurrent setting based on pi-calculus there is a curry-howard correspondence of logic and session typed pi, where DF holds
\item talk about CP here
\item \cite{dardhagay18} brought the research fwd and added priorities a la Kobayashi and Padovani to CP, producing PCP
\item following that lead, we present PGV
\item running example of the scheduler in PGV (hopefully)
\item translation of PCP to PGV and what it means to have that.
\item a contribution paragraph with the following
  \begin{itemize}
  \item PGV and what are the benefits
  \item updated PCP
  \item encoding PCP to PGV
  \item technical results to be referred to explicitly
  \end{itemize}
\end{itemize}

### 2020-10-09 priority-cp.tex
Responded to several TODOs, updated bibliography. Minor additions to introduction.
Cleaned up BibTeX files. Removed 'seminal'.

\begin{figure}[t]
  \centering
  \includegraphics[width=0.8\columnwidth]{scheduler}
  \vspace{-4mm}
  \caption{Cyclic Scheduler from \cite{dardha2018}}
  \label{fig:scheduler}
  \vspace{-4mm}
  \end{figure}

@@

We present an updated version of Priority CP~\cite[PCP]{dardha2018}, which is Wadler's Classical Processes~\cite[CP]{wadler2012} with \emph{priorities}.
The main difference with respect to PCP by Dardha and Gay \cite{dardha2018} is moving away from a reduction strategy based on cycle elimination using commuting conversions, and towards a reduction strategy using structural congruence.

@@

The types $\ty{\typlus[\cs{o}]{A}{B}}$ and $\ty{\tywith[\cs{o}]{A}{B}}$ type the endpoints of a channel over which we can receive or send a choice between two branches $\ty{A}$ or $\ty{B}$. We have opted for a simplified version of choice and followed the original Wadler's CP \cite{wadler2012}, however types $\oplus$ and $\with$ can be trivially generalised to $\ty{\oplus^{\cs{o}}\{l_i:A_i\}_{i\in I}}$ and $\ty{\with^{\cs{o}}\{l_i:A_i\}_{i\in I}}$, respectively as in the original PCP \cite{dardha2018}.
The types $\ty{\tynil[\cs{o}]}$ and $\ty{\tytop[\cs{o}]}$ type the endpoints of a channel over which we can send or receive a choice between \emph{no options}. These two types act as units for $\ty{\typlus[\cs{o}]{A}{B}}$ and $\ty{\tywith[\cs{o}]{A}{B}}$, respectively.

@@

Typing rules make use of priorities, defined in the previous section. Priorities are based on \emph{obligations/capabilities} used by Kobayashi \cite{kobayashi2006}, and simplified to single priorities following Padovani \cite{padovani2014}. Recalling PCP \cite{dardha2018}, priorities obey the following two laws:
\begin{enumerate}
  \item [(i)] an action of priority $\cs o$ must be prefixed only by actions of priorities \emph{strictly smaller} than $\cs o$.
  \item [(ii)] communication requires \emph{equal} priorities for the dual actions.
\end{enumerate}
\textsc{T-Link} states that the link process $\tm{\link{x}{y}}$ is well typed under channels $\tm x$ and $\tm y$ having dual types, respectively $\ty A$ and $\ty {\co A}$. \textsc{T-Res} sates that the restriction process $\tm{\res{x}{y}{P}}$ is well typed under typing context $\ty\Gamma$ if process $\tm P$ is well typed in $\ty\Gamma$ augmented with channel endpoints $\tm x$ and $\tm y$ having dual types, respectively $\ty A$ and $\ty {\co A}$. \textsc{T-Par} states that the parallel composition of processes $\tm P$ and $\tm Q$ is well typed under the disjoint union of their respective typing contexts. \textsc{T-Halt} states that the terminated process $\tm{\halt}$ is well typed in the empty context.

@@

\begin{example}[Milner's Cyclic Scheduler \Cref{fig:scheduler}]
  % 
  We will illustrate PCP with the cyclic scheduler \cite{milner99}. The full typing for $\tm {Sched}$ is given in \cite{dardha2018}.

@@

The reduction relation is given by a set of axioms and inference rules for context closure. Reduction occurs under restriction following \cite{dardha2018} where channel endpoints are identified.
$\LabTirName{E-Link}$ reduces a parallel composition with a link into a substitution. $\LabTirName{E-Send}$ is the main communication rule, where send and receive processes sychronise and reduce to the corresponding continuations. $\LabTirName{E-Close}$ follows the previous rule and it closes the channel identified by endpoints $\tm x$ and $\tm y$. $\LabTirName{E-Select-Inl}$ and $\LabTirName{E-Select-Inr}$ are generalised versions of $\LabTirName{E-Send}$. They state respectively that a left and right selection synchronises with a choice offering and reduces to the corresponding continuations. The last three rules state that reduction is closed under restriction, parallel composition and structural congruence, respectively.

@@

\begin{restatablelemma}{lempcpsubjectcongruence}[Subject Congruence, $\equiv$]
  \label{lem:pcp-subject-congruence}
  \hfill\\%newline before theorem statement
  If $\seq{P}{\ty{\Gamma}}$ and $\tm{P}\equiv\tm{Q}$,
  then $\seq{Q}{\ty{\Gamma}}$.
\end{restatablelemma}
\begin{proof}
  See~\cite{dardha2018}.
\end{proof}

\begin{restatabletheorem}{thmpcpsubjectreduction}[Subject Reduction, $\red$]
  \label{thm:pcp-subject-reduction}
  \hfill\\%newline before theorem statement
  If $\seq{P}{\ty{\Gamma}}$ and $\tm{P}\red\tm{Q}$, then $\seq{Q}{\ty{\Gamma}}$.
\end{restatabletheorem}
\begin{proof}
  See~\cite{dardha2018}. \todo{SR in \cite{dardha2018} uses commuting conversions. I guess here the result follows as we are simply considering a subset of reductions?}
\end{proof}

\begin{definition}[Actions]
  A~process acts on an endpoint $\tm{x}$ if it is in one of the following forms:
  \begin{multicols}{3}
    \begin{itemize}[noitemsep,topsep=0pt,parsep=0pt,partopsep=0pt]
    \item $\tm{\link{x}{y}}$ 
    \item $\tm{\link{y}{x}}$
    \item $\tm{\send{x}{y}{P}}$
    \item $\tm{\recv{x}{y}{P}}$
    \item $\tm{\close{x}{P}}$
    \item $\tm{\wait{x}{P}}$
    \item $\tm{\inl{x}{P}}$
    \item $\tm{\inr{x}{P}}$
    \item $\tm{\offer{x}{P}{Q}}$
    \item $\tm{\absurd{x}}$
    \end{itemize}
  \end{multicols}
  \noindent
  A~process is an action if it acts on some endpoint $\tm{x}$.
  \todo{Can't we simplify this def and say that every process is an action except restriction, par and zero.}
\end{definition}

>>>>

\begin{figure}[t]
  \centering
  \includegraphics[width=0.8\columnwidth]{scheduler}
  \vspace{-4mm}
  \caption{Cyclic Scheduler from \cite{dardhagay18}}
  \label{fig:scheduler}
  \vspace{-4mm}
  \end{figure}

@@

We present an updated version of Priority CP~\cite[PCP]{dardhagay18}, which is Wadler's Classical Processes~\cite[CP]{wadler12} with \emph{priorities}.
The main difference with respect to PCP by Dardha and Gay \cite{dardhagay18} is moving away from a reduction strategy based on cycle elimination using commuting conversions, and towards a reduction strategy using structural congruence.

@@

The types $\ty{\typlus[\cs{o}]{A}{B}}$ and $\ty{\tywith[\cs{o}]{A}{B}}$ type the endpoints of a channel over which we can receive or send a choice between two branches $\ty{A}$ or $\ty{B}$. We have opted for a simplified version of choice and followed the original Wadler's CP \cite{wadler12}, however types $\oplus$ and $\with$ can be trivially generalised to $\ty{\oplus^{\cs{o}}\{l_i:A_i\}_{i\in I}}$ and $\ty{\with^{\cs{o}}\{l_i:A_i\}_{i\in I}}$, respectively as in the original PCP \cite{dardhagay18}.
The types $\ty{\tynil[\cs{o}]}$ and $\ty{\tytop[\cs{o}]}$ type the endpoints of a channel over which we can send or receive a choice between \emph{no options}. These two types act as units for $\ty{\typlus[\cs{o}]{A}{B}}$ and $\ty{\tywith[\cs{o}]{A}{B}}$, respectively.

@@

Typing rules make use of priorities, defined in the previous section. Priorities are based on \emph{obligations/capabilities} used by Kobayashi~\cite{kobayashi06}, and simplified to single priorities following Padovani~\cite{padovani14}. Recalling PCP \cite{dardhagay18}, priorities obey the following two laws:
\begin{enumerate}
  \item [(i)] an action of priority $\cs o$ must be prefixed only by actions of priorities \emph{strictly smaller} than $\cs o$.
  \item [(ii)] communication requires \emph{equal} priorities for the dual actions.
\end{enumerate}
\textsc{T-Link} states that the link process $\tm{\link{x}{y}}$ is well typed under channels $\tm x$ and $\tm y$ having dual types, respectively $\ty A$ and $\ty {\co A}$. \textsc{T-Res} sates that the restriction process $\tm{\res{x}{y}{P}}$ is well typed under typing context $\ty\Gamma$ if process $\tm P$ is well typed in $\ty\Gamma$ augmented with channel endpoints $\tm x$ and $\tm y$ having dual types, respectively $\ty A$ and $\ty {\co A}$. \textsc{T-Par} states that the parallel composition of processes $\tm P$ and $\tm Q$ is well typed under the disjoint union of their respective typing contexts. \textsc{T-Halt} states that the terminated process $\tm{\halt}$ is well typed in the empty context.

@@

\begin{example}[Milner's Cyclic Scheduler \Cref{fig:scheduler}]
  % 
  We will illustrate PCP with the cyclic scheduler~\cite{milner89}. The full typing for $\tm {Sched}$ is given in \cite{dardhagay18}.

@@

The reduction relation is given by a set of axioms and inference rules for context closure. Reduction occurs under restriction following \cite{dardhagay18} where channel endpoints are identified.
$\LabTirName{E-Link}$ reduces a parallel composition with a link into a substitution. $\LabTirName{E-Send}$ is the main communication rule, where send and receive processes sychronise and reduce to the corresponding continuations. $\LabTirName{E-Close}$ follows the previous rule and it closes the channel identified by endpoints $\tm x$ and $\tm y$. $\LabTirName{E-Select-Inl}$ and $\LabTirName{E-Select-Inr}$ are generalised versions of $\LabTirName{E-Send}$. They state respectively that a left and right selection synchronises with a choice offering and reduces to the corresponding continuations. The last three rules state that reduction is closed under restriction, parallel composition and structural congruence, respectively.

@@

\begin{restatablelemma}{lempcpsubjectcongruence}[Subject Congruence, $\equiv$]
  \label{lem:pcp-subject-congruence}
  \hfill\\%newline before theorem statement
  If $\seq{P}{\ty{\Gamma}}$ and $\tm{P}\equiv\tm{Q}$,
  then $\seq{Q}{\ty{\Gamma}}$.
\end{restatablelemma}
\begin{proof}
  See~\cite{dardhagay18}.
\end{proof}

\begin{restatabletheorem}{thmpcpsubjectreduction}[Subject Reduction, $\red$]
  \label{thm:pcp-subject-reduction}
  \hfill\\%newline before theorem statement
  If $\seq{P}{\ty{\Gamma}}$ and $\tm{P}\red\tm{Q}$, then $\seq{Q}{\ty{\Gamma}}$.
\end{restatabletheorem}
\begin{proof}
  See~\cite{dardhagay18}.
\end{proof}

\begin{definition}[Actions]
  A~process acts on an endpoint $\tm{x}$ if it is in one of the following forms:
  \begin{multicols}{3}
    \begin{itemize}[noitemsep,topsep=0pt,parsep=0pt,partopsep=0pt]
    \item $\tm{\link{x}{y}}$ 
    \item $\tm{\link{y}{x}}$
    \item $\tm{\send{x}{y}{P}}$
    \item $\tm{\recv{x}{y}{P}}$
    \item $\tm{\close{x}{P}}$
    \item $\tm{\wait{x}{P}}$
    \item $\tm{\inl{x}{P}}$
    \item $\tm{\inr{x}{P}}$
    \item $\tm{\offer{x}{P}{Q}}$
    \item $\tm{\absurd{x}}$
    \end{itemize}
  \end{multicols}
  \noindent
  A~process is an action if it acts on some endpoint $\tm{x}$.
\end{definition}
